Practical Optional Types for Clojure

### 2015-07-10 formal-model.tex
add LOC. add HMap defalias counts. explain zero MM in f2imap
initial POPL submission

\subsection{Core type system}

@@

Next, we walk through checking
\begin{equation}  \label{eq:hi-defmethoden}
{\extendmultiexp {hi} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
\end{equation}

@@

The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
we only model the former to avoid the added complexity of keywords being functions.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}},
usually written {\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.

@@

The typing rules for \getliteral{} consider three possible cases. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
Lookups on unions of HMaps are only supported in T-GetHMap, 
for cases like \egref{example:desserts-on-meal}
where every element in the union
contains the key we are looking up.
The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
is a symbolic representation for a keyword lookup of $k$ in \object{}.
The substitution for {\x{}} handles the case where \object{} is empty.
$$
\begin{array}{rcl}
{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
\end{array}
$$

@@

% checks obvious prop logic, but update is most interesting rule.

>>>>

\subsection{Core type system}
\label{sec:coretypesystem}

@@

Next, we walk through checking
\begin{equation}  \label{eq:hi-defmethoden}
{\extendmultiexp {hi_0} {\makekw{en}} {\abs {\x{}} {\Keyword} {\makestr{hello}}}}
\end{equation}

@@

The expressions \clj{(get m :a)} and \clj{(:a m)} are semantically identical, though
we only model the former to avoid the added complexity of keywords being functions.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps %{\curlymapvaloverright{\k{}}{\v{}}},
%usually 
written 
{\curlymapvaloverrightnoarrow{\k{}}{\v{}}}.

@@

The typing rules for \getliteral{} consider three cases. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
Lookups on unions of HMaps are only supported in T-GetHMap, 
for cases like \egref{example:desserts-on-meal}
where every element in the union
contains the key we are looking up.
The object {\replacefor {\path {\keype{k}} {\x{}}} {\object{}} {\x{}}}
is a symbolic representation for a keyword lookup of $k$ in \object{}.
The substitution for {\x{}} handles the case where \object{} is empty.
$$
\begin{array}{rcl}
{\replacefor {\path {\keype{k}} {\x{}}} {\y{}} {\x{}}} &=& {\path {\keype{k}} {\y{}}} \\
{\replacefor {\path {\keype{k}} {\x{}}} {\emptyobject{}} {\x{}}} &=& \emptyobject{}
\end{array}
$$

@@

% checks obvious prop logic, but update is most interesting rule.
Now that we have covered \lambdatc{}, we now review
the proof system, in particular the T-Update rule.
The rest of the proof system, which uses standard proposition logic, 
we defer to the supplemental material---\figref{appendix:figure:proofsystem} contains the full rules.
$$
\LUpdate{}
$$
The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).
The rule says if we know positive information \t{} about object \path{\pathelemp{}}{\x{}}, and some
either positive or negative information \propisnotmeta{} about some extension of that path,
\path{\pathelem{}}{\path{\pathelemp{}}{\x{}}}, 
the \updateliteral{} metafunction can derive some new positive information about \path{\pathelemp{}}{\x{}}.

The example in 
\secref{sec:coretypesystem}
that combines information
$$
{ \inpropenv 
  {{\isprop{\Order}{\makelocal{o}}},
    {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}}}
  {\isprop {\t{}} {\makelocal{o}}}}
$$
to deduce \t{},
uses T-Update where \path{\pathelemp{}}{\x{}} = {\path{\keype{\makekw{Meal}}}{\makelocal{o}}}
and
$$
\update{\Order}{\makekw{combo}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}
$$
Since {\Order} is just a union of HMaps, we update each member of the union
via the last HMap case of
\figref{main:figure:update}.
Each recursive call hits the first HMap case, since {\makekw{combo}}
is always present, finally calling \restrictliteral{} in the penultimate case.

The non-combo HMaps change to \Bot, because \restrict{\makekw{lunch}}{\makekw{combo}} = \Bot,
leaving just the \makekw{combo} HMap.
